|

### 2021-04-12 04-Chp-Systematic-Review.Rmd
Test

* Smeet et al. reference in the CPRD analysis

@@

>>>>

* Smeet et al. reference in the CPRD analysis
* Larsson - 10.1136/bmj.j5375 [@larsson2017b]

@@

### Searching for/including Mendelian randomisation studies

As a form of analysis that lends itself to multiple comparison, there is a strong

Through snowballing and other measurse, we identified [@larsson2017b]

Is it a big issue that the studies have the same underlying datasets? <!--- see discussion from Matt Lee thesis, in email --->

### 2021-04-12 tmp.Rmd
Test

Chapter 1. Introduction
1.1. Summary
This chapter presents the problem that the research conducted in this thesis seeks to
address and an outline of the main output of this thesis: new evidence concerning
repurposing antihypertensives for dementia prevention using genetic and non-genetic
instrumental variable analyses. This chapter then introduces three issues that arose and
have been addressed during the conduct of this research. They are: how to calculate
power for instrumental variable analyses in pharmacoepidemiology; how to use
Mendelian randomization to predict drug repurposing opportunities; and what factors
have effected prescribing of existing dementia drugs. This chapter concludes with the
aims and objectives of this thesis, a summary of its organization and a list of the
associated outputs.
1.2. Statement of the problem
There is a substantial unmet clinical need for treatments for dementia where significant
benefits to patients, society and the public purse can be gained. Despite this, some drug
companies have recently withdrawn from this therapy area due to failed and costly efforts
to find new treatments. (1,2) Drug repurposing, the identification of properties in existing
or abandoned compounds for other clinical conditions, offers significant advantages over
traditional drug discovery approaches. This includes immediate access to human safety
data from the original clinical development work, which can accelerate testing in clinical
trials, saving both time and money. (3–5)
Many antihypertensive medications have been proposed as drug repurposing candidates
for the prevention of dementia. In part, because of research to better understand several
reports of observed associations between midlife hypertension and later-life risk of
Alzheimer’s disease and vascular dementia. (5–8) There is also increasing recognition
that one of the earliest pathological events in the development of Alzheimer’s disease is
vascular dysregulation. (9) As well as suggestions that some antihypertensives,
specifically those that block angiotensin receptor and calcium channel signalling, may
have other neurological benefits. (9–11)
2
Several observational studies have investigated repurposing antihypertensives for
dementia prevention. (12–19) However, these studies have typically used case-control
designs with logistic regression and cohort designs with survival analysis, both of which
are observational study designs that may be subject to unmeasured or residual
confounding and reverse causation. Specific concerns of observational studies are
confounding by indication, where the reasons that a patient receives a treatment relate to
the reasons that the patient is at an increased risk of the outcome; and healthy adherer
bias, where patients initiating or adhering to a drug for prevention of a condition are
more likely to be healthy. There is also potential for reverse causation due to preclinical
or early stages of the disease, which could lead to more frequent contacts with general
practitioners (GPs) and lead to raised blood pressure being more likely to be detected.
This, in turn, could lead to the prescription of an antihypertensive drug in advance of
dementia being formally diagnosed. Consequently, current evidence concerning
repurposing antihypertensives for dementia prevention is considered inconclusive.
1.3. New evidence concerning repurposing antihypertensives for dementia prevention
using instrumental variable analysis
Instrumental variable analysis, which estimates the causal effect of an exposure on an
outcome by using a third variable (the instrument), can be robust to confounding and
reverse causation if certain assumptions are met. That is:
  IV1. The instrument must associate with the exposure
IV2. The instrument must only affect the outcome through the exposure
IV3. The instrument and the outcome must have no common causes
These assumptions can be represented on a directed acyclic graph, as shown in Figure
1.1. Instrumental variable analysis, with other sources of evidence, can be used in a
triangulation framework to obtain a reliable answer concerning the potential repurposing
of antihypertensives for dementia prevention. (20) This thesis presents two forms of
instrumental variable analysis to provide new evidence concerning this hypothesis. The
first uses physicians’ prescribing preference as a non-genetic instrument in electronic
health record data obtained from the United Kingdom (UK) Clinical Practice Research
Datalink (CPRD). The second uses single nucleotide polymorphisms (SNPs) i.e.
differences in the deoxyribonucleic acid (DNA) nucleotides between individuals, which
have been selected to mimic the biological function of the protein targets of 
3
antihypertensive drug classes, as a genetic instrument in an approach more commonly
known as Mendelian randomization. The results of both these instrumental variable
analyses have been made available via the following references. (21,22)
1.4. Calculating power for instrumental variable analysis in pharmacoepidemiology
Instrumental variable analysis is an increasingly popular method in the field of
pharmacoepidemiology. (23–28) However, the power calculators that were available for
studies using instrumental variable analysis at the start of my PhD – such as Mendelian
randomisation power calculators – did not allow for the structure of research questions
using non-genetic instruments (for example, physicians’ prescribing preference) in this
field. (29,30) This is because analysis using non-genetic instruments in pharmacoepidemiology will typically have stronger instruments and so can detect smaller causal
effects. Consequently, there was a need for dedicated power calculators for these type of
research questions in pharmacoepidemiology. In this thesis, I investigate how to conduct
Figure 1.1: A directed acyclic graph illustrating the basic instrumental variable analysis model.
Directed acyclic graphs are a visual representation of a model, which represent variables by ‘nodes’
and the relationships between them by ‘directed edges’. (68) The graphs are defined as acyclic
because edges cannot form ‘cycles’, whereby the edges all act in the same direction. This prevents a
variable from being both a cause and a consequence of itself. Directed acyclic graphs are a common
tool in epidemiology and, more specifically, causal inference when edges are given a causal
interpretation, as they allow researchers to depict their model and its assumptions using a common
graphing scheme. (69)
4
power calculations for pharmacoepidemiological studies, which use a single binary
instrument to analyse the causal effect of a binary exposure on a continuous outcome. I
also provide an online calculator, as well as packages in both R and Stata, for the
implementation of the formula by others (https://github.com/venexia/PharmIV). This
work has been published in the International Journal of Epidemiology. (31)
1.5. Using Mendelian randomization to predict drug repurposing opportunities
Identification of drug repurposing opportunities can maximize the benefit of a drug.
However, as highlighted before, the more traditional observational research methods
used to investigate these opportunities are subject to several biases. These include
confounding by indication, reverse causality, and missing data. In this thesis, I propose
Mendelian randomization as a novel approach that can be used for the prediction of drug
repurposing opportunities. Mendelian randomization addresses some of the limitations
associated with the existing methods in this field. Furthermore, it can be applied either
pre- or post-approval of the drug and could therefore prevent the potentially harmful
exposure of patients in clinical trials and beyond. This thesis includes discussion of
examples from the literature that have used Mendelian randomization to predict drug
repurposing opportunities and covers the strengths and limitations associated with using
this method for this purpose. There was relatively little discussion focussed on using
Mendelian randomization for drug repurposing when I commenced my PhD and I have
since published on this topic. (32)
1.6. Factors effecting existing dementia drug prescribing
Drugs for dementia have been available in England from 1997. However, since their
launch, there have been several changes to national guidelines and initiatives that may
have influenced prescribing. These include changes in National Institute for Health and
Care Excellence (NICE) guidance; several government dementia strategies; the addition
of dementia to the Quality and Outcomes Framework (QOF); and the expiry of drug
patents. Despite this, little research had been conducted prior to my PhD into the effect
of these events on prescribing. (33,34) In this thesis, I investigate prescribing trends in
England since the launch of these drugs up to 1st January 2016 using data from the
CPRD to address this gap in the literature. The key motivation for this analysis was to 
5
identify factors that have affected prescriptions of existing treatments that may also
influence repurposed drug candidates in the future. However, the results from this
analysis could also be used to identify breaks in the prescription of these drugs that could
be exploited as natural experiments for progression studies. For example, if these drugs
were not prescribed when the NICE guidelines stopped recommending their use between
2006 and 2011, the progression of people diagnosed during this time who did not access
the drugs could be compared with the progression of people diagnosed before and after
this time who did have access to the drugs. The results presented in this chapter have
been published in Alzheimer’s Research & Therapy. (35)
1.7. Aims and objectives
The objective of this thesis was to use instrumental variable analysis methods, in existing
data sources, to triangulate evidence for repurposing antihypertensive drugs for the
prevention of dementia.
The specific aims were as follows:
  1. Develop a power calculator for non-genetic instrumental variable analysis studies
in the context of pharmacoepidemiology.
2. Describe the use of genetic instrumental variable analysis, namely Mendelian
randomization, for predicting drug repurposing opportunities.
3. Examine the impact of regulatory guidance and patent expiry on dementia drug
prescribing.
4. Investigate whether antihypertensive drugs have a causal effect on incident
dementia using instrumental variable analysis with electronic health record data.
5. Investigate whether antihypertensive drugs have a causal effect on incident
dementia using instrumental variable analysis with genetic data.
1.8. Organization of this thesis
Chapter 2 provides the background necessary for the rest of this thesis through the
introduction of dementia, the concept of drug repurposing, and discussion of the existing
evidence regarding antihypertensive drugs for dementia prevention. It also covers the 
6
strengths and limitations of observational pharmacoepidemiology and explains why it
might be preferable over ‘gold standard’ randomized controlled trials for some
hypotheses. Chapter 3 describes instrumental variable analysis, the method utilized
throughout this thesis, and documents the development of a power calculator for this
method in the context of pharmacoepidemiology (Aim 1). Chapter 4 introduces the idea
of using Mendelian randomization, a form of instrumental variable analysis that uses
genetic variants as instruments, for drug repurposing and covers the strengths and
limitations of this approach (Aim 2). Chapter 5 describes the CPRD and my use of this
data source. Chapter 6 covers the current treatments available for dementia and factors
affecting their prescription in England based on data from the CPRD (Aim 3). Chapter 7
presents an assessment of the effects of antihypertensive drugs on dementia prevention
using instrumental variable analysis with data from the CPRD (Aim 4); while Chapter 8
presents an assessment of the effects of antihypertensive drugs on dementia prevention
using instrumental variable analysis with genetic data (Aim 5). The thesis concludes with
a discussion in Chapter 9 that brings together all the elements of this thesis and discusses
their implications.
1.9. Outputs from this thesis
1.9.1. Contributions to scientific literature
Contributions to scientific literature arising from this thesis are detailed below and
provided in Appendix A.
The protocol for the observational work using CPRD data to investigate drug
repurposing opportunities is published in the BMJ Open. (36) Its contents are referenced
in several places throughout this thesis, particularly Chapter 5:
  Walker VM, Davies NM, Jones T, Kehoe PG, Martin RM. Can commonly prescribed
drugs be repurposed for the prevention or treatment of Alzheimer’s and other
neurodegenerative diseases? Protocol for an observational cohort study in the UK
Clinical Practice Research Datalink. BMJ Open. 2016 Dec 1;6(12):e012044.
7
The power calculator for instrumental variable analysis in pharmacoepidemiology,
described in Chapter 3, is published in the International Journal of Epidemiology (31):
  Walker VM, Davies NM, Windmeijer F, Burgess S, Martin RM. Power calculator for
instrumental variable analysis in pharmacoepidemiology. Int J Epidemiol. 2017 Oct
1;46(5):1627–32.
Also published in the International Journal of Epidemiology is an article discussing
Mendelian randomization as a novel approach for the prediction of adverse drug events
and drug repurposing opportunities (32), which formed the basis of Chapter 4:
  Walker VM, Davey Smith G, Davies NM, Martin RM. Mendelian randomization: a
novel approach for the prediction of adverse drug events and drug repurposing
opportunities. Int J Epidemiol. 2017 Dec 1;46(6):2078–89.
The trend analysis examining prescribing practice for drugs for dementia in the CPRD,
presented in Chapter 6, is available from Alzheimer’s Research & Therapy (35):
  Walker VM, Davies NM, Kehoe PG, Martin RM. What is the impact of regulatory
guidance and expiry of drug patents on dementia drug prescriptions in England? A trend
analysis in the Clinical Practice Research Datalink. Alzheimer’s Research & Therapy.
2018 May 29;10:51.
The assessment of antihypertensives for dementia prevention using electronic health
record data, reported in Chapter 7, is currently under peer review and available from
bioRxiv (21):
  Walker VM, Davies NM, Martin RM, Kehoe PG. Comparison of antihypertensive drug
classes for dementia prevention. bioRxiv. 2019 Jan 12;517482.
The assessment of antihypertensives for dementia prevention using genetic data, reported
in Chapter 8, has been made available as an advance article from the International
Journal of Epidemiology (22):
  8
Walker VM, Kehoe PG, Martin RM, Davies NM. Repurposing antihypertensive drugs
for the prevention of Alzheimer’s disease: a Mendelian Randomization study. Int J
Epidemiol. 2019 Jul 4; Advance article.
1.9.2. Contributions to scientific meetings
I presented the paper “Power calculator for instrumental variable analysis in
pharmacoepidemiology”, described in Chapter 3, at the UK Administrative Data
Research Network Annual Research Conference 2017 in Edinburgh, UK.
I presented the paper “Mendelian randomization: a novel approach for the prediction of
adverse drug events and drug repurposing opportunities”, which formed the basis of
Chapter 4, at the University of Bristol Population Health Symposium 2016 in Bristol, UK
and at the International Society for Pharmacoepidemiology mid-year meeting 2018 in
Toronto, Canada.
I presented posters, based on the paper “What is the impact of regulatory guidance and
expiry of drug patents on dementia drug prescriptions in England?” detailed in Chapter
6, at the Alzheimer’s Research UK Research Conference 2016 in Aberdeen, UK and at
the University of Bristol brain research showcase and networking day 2018 in Bristol,
UK.
I presented “Can treatments for hypertension be repurposed for the treatment of
dementia?” at the Society of Epidemiologic Research annual conference 2018 in
Baltimore, United States of America (USA); at the European Congress of Epidemiology
2018 in Lyon, France and the International Society for Pharmacoepidemiology annual
meeting 2018 in Prague, Czech Republic. These presentations combined results from
Chapters 7 and 8.

>>>>

---
output: html_document
bibliography: bib/library.bib
csl: csl/nature.csl
space_betwee_paragraphs: true
fig_caption: true
always_allow_html: yes
link-citations: true
---

```{r include=FALSE}
# word count = 
doc.type <- knitr::opts_knit$get('rmarkdown.pandoc.to') # this will tell the .Rmd file what output you are knitting to (word, pdf, html) so that you cna use if else statements when making tables - html/pdf output tables do not work well in word. For word we need to use the 'flextable' packaged to make tables.
## packages
library(dplyr)
library(kableExtra)
library(flextable)
```

# Data visualisation {#chapter6}

<style>
body {
text-align: justify}
</style>

### *EpiViz: a tool to visualise large metabolomic association analyses* {-}

In Chapters \@ref(chapter4) and \@ref(chapter5) we observed large metabolic shifts as a result of increased adiposity. These results were presented using Circos plots. Production of the first stage of these plots was inefficient. This chapter details the development and implementation of a web application and `R` package that efficiently creates Circos plots that can be used by the community for similar analyses. The `R` package was used to re-make the plots presented in Chapters \@ref(chapter4) and \@ref(chapter5) and make those in subsequent chapters. <br>

\newpage

## Introduction
The metabolome (briefly discussed in Section \@ref(metabolites)) is a complex biological system comprised of the bodies entire small molecules. These small molecules, metabolites, take many forms and act in a multitude of pathways[@Griffin2006]. Metabolites are expressed differentially among cell types and carry information from genetic and non-genetic factors[@Johnson2016; @Wishart2019]. They are the ultimate end-point of a biological process and are therefore seen as the link between genotype and phenotype[@Griffin2006; @Su2014; @Chu2019; @Wishart2019]. <br>

There are potentially many thousands of metabolites[@Griffin2006] across different cell types, with expression associated with traits and diseases[@Guasch-Ferre2016; @Liesenfeld2013; @Johnson2016; @Moore2014; @Cirulli2019; @Wishart2018]. However, assigning individual metabolite changes to diseases is difficult[@Su2014; @Fearnley2016; @Chu2019] because of the complexity of their interrelationships. The metabolome is made up of many feedback and feed-forward loops, meaning metabolic products of one pathway will likely influence those of another. The intercorrelated nature of metabolites[@Rosato2018] is further complicated by the wide degree of genetic correlation[@Gallois2019] and shared genetic architecture[@Shin2014; @Kettunen2016; @Long2017; @Gallois2019]. Gaining global overview of association analyses will likely aid interpretation of results. Global overview is of particular importance in MR studies where robust selection of instrumental variables is required to meet methodological assumptions[@DaveySmith2003]. <br>

Forest plots are widely used in metabolomic association analyses[@May-Wilson2017; @Taylor2019; @Soares2019; @SantosFerreira2019; @Hartley2020; @Borges2017] to visualise and interpret results. They can be easily made, with a number of `R` packages, such as [`ggforestplot`](https://github.com/nightingalehealth/ggforestplot) and [`TwoSampleMR`](https://mrcieu.github.io/TwoSampleMR/#to-many-forest-plot)[@Hemani2018], available. However, when presenting more than 50 data points forest plots are cumbersome and many separate plots are needed. This makes gaining global overview difficult. <br>

An alternative approach is to compress information into a circular rather than vertical/horizontal form (see Figures 3 and 2 from [@Shin2014] and [@Kettunen2016] respectively). Circos plots[@Krzywinski2009], as implemented in many genetics studies, condense large amounts of information into informative visualisations (see Figure 1A from [@Saben2014]). However, Circos software is designed for genomic data and written in programming languages unfamiliar to many epidemiologists. The `R` package [`circlize`](https://jokergoo.github.io/circlize_book/book/index.html)[@Gu2014] provides most of the functionality of the Circos software but is designed for genomic information and requires an advanced understanding of `R`. Here I present `EpiViz`, a web application and `R` package, that builds on the [`circlize`](https://jokergoo.github.io/circlize_book/book/index.html)[@Gu2014] and [`ComplexHeatmap`](https://jokergoo.github.io/ComplexHeatmap-reference/book/)[@Gu2016] `R` packages. `EpiViz` enables epidemiologists with little programming experience to efficiently create Circos plots with hundreds of variables and gain global overview.

## Methods
### Circos plots for association analyses
`EpiViz` adapts and builds on the `Circlize` foundation to create Circos plots compatible with association analysis data. Circos plots are composed of six elements: template, plotting space, tracks, sections, data, legend (Figure \@ref(fig:annotated-circos-plot)). The template element is a square of defined proportions within which information is plotted. Each additional element is layered onto the template one after the other. The plotting space element is an empty circle which is layered and centred on top of the template. Data is plotted on to the plotting space. An optional extra of the Circos plot, the legend element takes the dimensions of the template and creates a separate plotting space that can be layered on to the bottom of the template element. <br>

The plotting space is separated into tracks and sections. Tracks are laid down as rings within the plotting space. Each track represents a single element of information such as an exposure. Tracks are numbered from the outside to the centre of the circle and coloured separately. Sections divide the plotting space into distinct areas, much like a pie chart. Sections are defined by the data and usually represent groups of outcomes such as metabolite classes. A section track is laid at the outside of the tracks to give a header for each section. The header is referenced in the legend element. <br>

Once the template, plotting space, tracks and sections are laid down, coordinates for each section and track location can be called to plot the data element. Each track and section coordinate, e.g. track 2 section 3, is treated as an individual plotting space. As such, data can be plotted based on the following coordinates: track, section, $X$, $Y$. The $X$ axis of each track is defined by the number of rows in the data frame, i.e a data frame with 100 rows will have an $X$ axis of length 100 with each row given an $X$ axis coordinate from 1:100. The $Y$ axis is defined by the minimum and maximum of the data for that track. As such, each track and section coordinate, e.g. track 2 section 3, can be considered an individual plot with a $Y$ axis that is shared by all of the sections in that track. For each position on the $X$ axis the label element of each row is plotted outside of the section header. <br>

```{r annotated-circos-plot, echo=FALSE, out.width='100%', fig.cap="Annotated Circos plot highlighting elements of the Circos plot"}
knitr::include_graphics("data/chapter6/figures/annotated_circos_plot.pdf")
```

### Implementation
`EpiViz` is a `Shiny` web application and `R` package. `R` (version 3.6.2)[@r2019] and `Shiny`[@Chang2019] (version 1.4.0) were used to develop the web application and `R` (version 3.6.2) was used to develop the `R` package. `Shiny` is an `R` package that enables development and deployment of web applications written in the `R` programming language. Development of `EpiViz` was progressive and feedback from colleagues (see Section \@ref(epiviz-acknowledgements)) was vital in this process.

### Operation
The [web application](http://mrcieu.mrsoftware.org/EpiViz/) is publicly accessible -- [http://mrcieu.mrsoftware.org/EpiViz/](http://mrcieu.mrsoftware.org/EpiViz/) -- and held under an [MIT license](https://github.com/mattlee821/EpiViz/blob/master/LICENSE.txt). [Docker](https://www.docker.com/) was used to containerise the application. The [Medical Research Council Integrative Epidemiology Unit](http://mrcieu.mrsoftware.org/EpiViz/) at the University of Bristol `Shiny` server hosts the web application. The web application has been tested on computers running macOS (version 10.14) and Windows (version 10) using: Internet Explorer (version 11; Windows), Google Chrome (version 79; macOS and Windows), Safari (version 13; macOS). <br>

The [`R` package](https://github.com/mattlee821/EpiViz/tree/master/R_package) is publicly accessible through GitHub -- https://github.com/mattlee821/EpiViz/tree/master/R_package --  and held under an[MIT license](https://github.com/mattlee821/EpiViz/blob/master/LICENSE.txt). The `R` package is accessible on all computers with `R` version 3.3.0 or higher and has been tested on macOS (version 10.14) and Windows (version 10) running `R` version 3.3.0 or higher. <br>

A legend function is available for both the web application and `R` package and is implemented using functions from the [`ComplexHeatmap`](https://jokergoo.github.io/ComplexHeatmap-reference/book/)[@Gu2016] `R` package. By default the colours used for the Circos plot in both the web-application and `R` package are accessible colours identified using [i want hue](https://medialab.github.io/iwanthue/). Example data is provided on the web application *Home* tab and with the `R` package using the function `EpiViz::EpiViz_data*()` where `*` is 1-3. Example data can be produced for use with the web-application and `R` package using code on [GitHub](https://github.com/mattlee821/EpiViz/blob/master/R_package/data-raw/example_data_simulation.R). <br>

#### `Shiny` web application
In order to use the [web application](http://mrcieu.mrsoftware.org/EpiViz/) a web-browser and an internet connection of at least 1Mbps is required. No other system requirements are needed. Upon opening the web application, users are shown a number of example Circos plots created with the application and are directed towards the [*Home*]() tab. The *Home* tab (Figure \@ref(fig:web-app-home)) provides users with a short summary of the application, link to the `R` package, and example data for use with the app. <br>

```{r web-app-home, echo=FALSE, out.width='100%', fig.cap="Screenshot of the EpiViz web application Home page"}
knitr::include_graphics("data/chapter6/figures/web_app_home.png")
```

The [*How to*]() provides instructions for using the application and is split into three: Step 1 deals with the preparation of data to be used with the application, step 2 deals with how to use the application, and step 3 provides information on potential next steps. Users are instructed to upload one data frame per track of the Circos plot. Each data frame should be a tab delimited text file and `R` code is provided for users to achieve this. The user is guided through an example utilising the example data provided with the application and downloadable on the *Home* tab. <br>

Having prepared their data as instructed in Step 1, users select the [*Analysis*]() tab and upload one data frame for each track of the Circos plot. If formatted as described in the *How to* tab, descriptive information including a volcano plot will be produced automatically (Figure \@ref(fig:web-app-analysis-step1)). <br>

```{r web-app-analysis-step1, echo=FALSE, out.width='100%', fig.cap="Screenshot of the EpiViz web application Analysis page, showing Step 1"}
knitr::include_graphics("data/chapter6/figures/web_app_analysis_data.png")
```

If the data frame appears as expected to the user the Circos plot can then be produced by selecting the [*Plot*]() tab. A series of drop down lists are presented which auto-populate with the column names from the first uploaded data frame. The following column information is required for the Circos plot: label, group, estimate, p-value, lower confidence interval, upper confidence interval (Figure \@ref(fig:web-app-analysis-step2)). Users can select one, two, or three tracks for their Circos plot, with data for the respective track coming from the upload data section. <br>

```{r web-app-analysis-step2, echo=FALSE, out.width='100%', fig.cap="Screenshot of the EpiViz web application Analysis page, showing Step 2"}
knitr::include_graphics("data/chapter6/figures/web_app_analysis_plot.png")
```

A p-value threshold (default 0.05) can be provided to indicate a significance threshold. On the Circos plot a solid point is indicated as reaching the significance threshold and an open point indicates not reaching the threshold. An optional legend function is provided and users can choose the labels for the legend points. The legend is auto-populated using the levels in the `group` column of the up-loaded data frame. Finally, users can select to use *Accessible colours* (default) or *Not accessible colours* for the colours of the points on the Circos plot. <br>

#### `R` package
The `EpiViz` `R` package is accessed using `R` version 3.3.0 or above. Documentation for using the package to create Circos plots is available as a README on [GitHub](https://github.com/mattlee821/EpiViz/tree/master/R_package) and includes use cases and troubleshooting. The `R` package can be installed directly from GitHub into `R` with the following `R` code: <br>

```{r eval=FALSE}
# Install devtools
install.packages("devtools")
library(devtools)

# Install directly from GitHub
devtools::install_github("mattlee821/EpiViz/R_package")
library(EpiViz)
```

Once installed, users can use the example data provided with the package to produce Figure \@ref(fig:r-package-3-track-type-plot), which illustrates the three different types of track available (point, line, bar), using the following `R` code: <br>

```{r eval=FALSE}
circos_plot(track_number = 3,
            track1_data = EpiViz::EpiViz_data1,
            track2_data = EpiViz::EpiViz_data2,
            track3_data = EpiViz::EpiViz_data3,
            track1_type = "points",
            track2_type = "lines",
            track3_type = "bar",
            label_column = 1,
            section_column = 9,
            estimate_column = 2,
            pvalue_column = 3,
            pvalue_adjustment = 1,
            lower_ci = 4,
            upper_ci = 5,
            lines_column = 2,
            lines_type = "o",
            bar_column = 2)
```

```{r r-package-3-track-type-plot, echo=FALSE, out.width='100%', fig.cap="Circos plot prduced using EpiViz R package and example data"}
knitr::include_graphics("data/chapter6/figures/r_package_3_track_type_plot.pdf")
```

These Circos plots are designed to be viewed on a screen rather than on paper. As such the plots are created at a larger scale compared to traditional plots made in `R`. This means that when viewing the Circos plot in an viewer such as the `R` Studio Viewer pane the plot appears squashed. Viewers like this should be used only as a guide when making the plot. Instead, plots should saved as `PDF` using the below code -- `PDF` files can be converted to other image formats. <br>

```{r eval=FALSE}
pdf("my_circos_plot.pdf",
    width = 30, height = 30, pointsize = 35)
circos_plot(...
            circle_size = 25)
dev.off()
```

`R` sets default height, width and pointsize parameters and users should iterate the sizing of the plot adjusting the `width` and `height` functions to get the desired plot size. They should then adjust the `pointsize` function to increase and decrease the size of all information in the plot (labels, points, lines etc.). The numbers provided work with most plots and were used for all examples. In addition, users can adjust the size of the Circos plot directly using the argument `circle_size`. The default for `circle_size` is 25 and smaller numbers increase the size of the circle and larger numbers decrease the size. Finally, users can adjust the height of tracks individually using `track*_height` where `*` refers to the track number (see below code). The default for each track is 20 percent of the total size of the circle (the section track is fixed at 5 percent). <br>

```{r eval=FALSE}
circos_plot(...
            track1_height = 0.2,
            track2_height = 0.2,
            track3_height = 0.2)
```

In order to minimise the time required to maintain the `R` package further customisation is achieved by the users themselves. The function is written to aid this customisation with *Default parameters* and *Customisable parameters* located at the top of the `circos_plot()` function script. Guidance and code is provided on [GitHub](https://github.com/mattlee821/EpiViz/tree/master/R_package) to aid user customisation, for example changing track colours requires users navigating to the *Customisable parameters* section and then to the *Colours* section and replacing the following lines with their desired colours: <br>

```{r eval=FALSE}
discrete_palette <- c("#00378f", # track 1 colour
                      "#ffc067", # track 2 colour
                      "#894300") # track 3 colour
```

## Discussion
Circos plots have been used to visualise and provide overview of large association analyses both in the literature[@Shin2014; @Kettunen2016] and this thesis (see Chapters \@ref(chapter4), \@ref(chapter5) and \@ref(chapter9)). This has been time consuming and cumbersome. Requiring bespoke code these plots have been challenging to work with when using multiple data sets (see [GitHub](https://github.com/mattlee821/000_thesis/tree/master/index/data/chapter6/code/circos_example.R) for example bespoke code). <br>

`EpiViz` provides an efficient way to produce Circos plots for association analyses and can be implemented with little up-front cost. The web application is intended for researchers with limited to no experience of `R`, while those with some experience can gain added functionality from the `R` package. The application of Circos plots to comparing and gaining global overview of association analyses is presented in Chapters \@ref(chapter4), \@ref(chapter5) and \@ref(chapter9), as well as by colleagues[@Taylor2019]. <br>

Though intended to be low maintenance, there is scope for further development of both the web application and `R` package. Of broad appeal to both is the inclusion of new plotting methods such as the chord diagram which shows relatedness between two nodes. In the case of metabolites a chord diagram would provide greater understanding about the relationship between different sections. In addition, the ability to filter and choose specific sections to display would simplify the plot and highlight potentially meaningful results. <br>

The web application is less flexible than the `R` package, however there are specific extensions of interest. Firstly interactive plots would provide users the ability to expand sections of interest and display additional information. Rather than a final presentation of the data this would primarily focus on exploration of data. Additionally, introducing features from the `R` package such as different plot types will broaden the appeal of the web application. These additions are likely to increase the costs of maintenance however. <br>

The `R` package has extensive scope for enhancement and presents a more efficient investment of time for extensions over the web application due to the flexibility of `R`. Currently, only a single data point can be plotted on each track at each $X$ axis coordinate. This is to prevent over-crowding. However, if a data set has 50 points for example, adding a second point to each $X$ axis coordinate of the track would not overcrowd the plot. <br>

The current limited customisation afforded with the web application and the `R` package is designed to reduce the maintenance costs of the author. As a result, functions that users may wish to use, such as multiple point estimates on a single track, must be coded by the user in `R`. Though guidance is provided on GitHub for such customisations, this additional effort may reduce the utility of the tool, especially among users with little experience of `R` but very specific requirements. Similarly, although designed to be efficient, users must provide data in a specific format in order to use the web application and `R` package. `R` code is provided to aid this effort but the requirement for users with no `R` experience to create a specific data frame will be off-putting for some. <br>

In order to achieve the desired goal of providing global overview of large association analyses Circos plots must be much larger than traditional plots such as forest plots. As such, Circos plots are best utilised as a tool for the user rather than a reader of a print journal. Circos plots are best viewed using a screen with the ability to zoom in to sections where more detail is required. This fact will reduce the utility of this type of Circos plot in publications as fixed margins will limit the size and resolution of the plot. Resolution can be improved by saving the Circos plots as a `PDF` file. Many journals do not accept `PDF` images and will convert these into lossless formats such as `PNG` or `JPEG`. This will reduce the resolution provided for readers. <br>

`EpiViz` is simple and efficient to use. There is potential for `EpiViz` to become widely used among the community. To pre-empt popularity and provide a stable environment the web application was containerized using Docker and hosted on a server with provision for multiple-users. With popularity it is hoped that a community will help to develop `EpiViz` further. <br>

## Acknowledgements {#epiviz-acknowledgements}
`EpiViz` was greatly aided by the `R` packages [`Circlize`](https://jokergoo.github.io/circlize_book/book/index.html)[@Gu2014] and [`ComplexHeatmap`](https://jokergoo.github.io/ComplexHeatmap-reference/book/)[@Gu2016], whose code is freely available and was used as a reference point. The following contributed to the development of `EpiViz`: Luke McGuinness, Osama Mahmoud, Sam Neaves. The following colleagues provided feedback in the development of `EpiViz`: Caroline Bull, Charlie Hatcher, Kurt Taylor, Nancy Mcbride, Neil Goulding, Steph Suddell.

\newpage
